Co-moving stars in \textsl{Gaia DR1}

### 2016-10-14 paper/ms.tex
still working on appendix

% packages
\definecolor{cbblue}{HTML}{3182bd}
\usepackage{microtype}  % ALWAYS!
\usepackage{amsmath}
\hypersetup{backref,breaklinks,colorlinks,urlcolor=cbblue,linkcolor=cbblue,citecolor=black}
\usepackage{dsfont}

@@

% define macros for math
\newcommand{\given}{\,|\,}
\newcommand{\normal}{{\mathcal{N}}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\transp}[1]{{#1}^{\!\mathsf{T}}}
\newcommand{\inv}[1]{{#1}^{-1}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\vperp}{\bs{v}^\perp}
\newcommand{\propm}{\bs{\mu}}
\newcommand{\matrx}[1]{\mathbf{#1}}
\newcommand{\kms}{\rm km~s^{-1}}
\newcommand{\msun}{{\rm M}_\odot}
\newcommand{\data}{\mathrm{data}}
\newcommand{\snr}{[S/N]_\varpi}
\newcommand{\eye}{\mathds{1}}

@@

For each star with $\snr > 32$, we use the reported parallax and parallax
error to compute a distance corrected for the Lutz-Kelker bias
(\citealt{lutzkelker}):
\begin{equation}
  d_{\rm LK} = 1000 \, \left[\frac{\varpi}{2} \,
    \left(1 + \sqrt{1 - \frac{16}{\snr^2}} \right) \right]^{-1} \, {\rm pc}
\end{equation}
We use this distance and the sky positions of each source to compute 3D
Cartesian positions, $\boldsymbol{x}$, for all stars and find the nearest
neighbor to each source in 3-space using the KDTree implementation in
\project{scikit-learn} (\citealt{scikit-learn}). We remove duplicates from
this sample and are left with 29870 pairs of stars. For all pairs, we then
compute the tangential velocity difference
\begin{equation}
  |\Delta \bs{v}_{\rm t}| = d_{\rm LK} \,
                            \sqrt{(\mu_{\alpha,1}-\mu_{\alpha,2})^2 +
                                  (\mu_{\delta,1}-\mu_{\delta,2})^2}
\end{equation}
\figname~\ref{fig:dv-sep} shows the tangential velocity difference against the
physical separation plotted for all pairs that pass the above signal-to-noise
cut. Pairs with small physical separation also tend to have small velocity
differences. Intriguingly, there appear to be an abundance of pairs with small
velocity separation at separations $|\delta\bs{x}| \gtrsim 1~{\rm pc}$. Of
course, this figure really only highlights close pairs with small angular
separations:
At large angular separations, the tangential velocity differences will be large
because of projection effects.
To select a more representative population of co-moving stars, we must instead
model the 3-space velocities of the stars given observations of just two
components.

@@

The abundance of pairs of stars with small velocity difference in the
high-precsision sample of \figname~\ref{fig:dv-sep} suggests that there are a
significant number of co-moving stars in the \tgas\ data at separations as large
as a few parsecs. Here we develop a method to select high-confidence co-moving
stars that properly incorporates the complex uncertainties associated with the
\gaia\ data. We make the following assumptions in order to construct a
statistical model (a likelihood function with explicit priors on our
parameters):
\begin{itemize}
  \item We assume that the uncertainties in the data---parallax, $\varpi$, and
    two proper motion components, $\propm = (\begin{array}[t]{c c} \mu_\alpha &
    \mu_\delta\end{array})^\mathsf{T}$---are Gaussian with known covariances
    $\matrx{C}$.
  \item We assume that the 3-space velocities of a given pair of stars
    $(\bs{v}_i, \bs{v}_j)$ in the \tgas\ sample (relative to the solar system
    barycenter) are either (1) co-moving with velocity drawn from a velocity
    prior $p(\bs{v})$ and  velocity difference drawn from a zero-mean Gaussian
    with velocity dispersion $s$, or (2) individually drawn from a velocity
    prior $p(\bs{v})$.
\end{itemize}

Under these assumptions, the likelihood of a proper motion measurement for a
star given true distance, $d$, and true tangential velocity, $\bs{v}^\perp =
(\begin{array}[t]{c c} v_\alpha & v_\delta\end{array})^\mathsf{T}$ is
\begin{align}
  L(\bs{v}, d, s^2) &=
    \left[\det\left(\frac{\tilde{\matrx{C}}^{-1}}{2\pi}\right)\right]^{1/2} \,
    \exp \left[ -\frac{1}{2} \transp{\left(\bs{\mu} - \bs{x}_\theta \right)} \,
    \tilde{\matrx{C}}^{-1} \,
    \left(\bs{\mu} - \bs{x}_\theta \right) \right] \\
  \bs{x}_\theta &= d^{-1} \, \bs{v}^\perp
\end{align}
where the tangential velocity $\bs{v}^\perp$ is related to the 3-space velocity
$\bs{v}$ through projection onto a tangent plane at a given star's sky position
$(\alpha, \delta)$
\begin{align}
  \bs{v}^\perp &= \matrx{M}\,\bs{v} \\
  & = \left(
      \begin{array}{c c c}
        -\sin\alpha & \cos\alpha & 0 \\
        -\sin\delta \, \cos\alpha & -\sin\delta \, \sin\alpha & \cos\delta
      \end{array}
    \right) \,
    \left(\begin{array}{c} v_x \\ v_y \\ v_z \end{array}\right)
\end{align}
and the modified covariance matrix $\tilde{\matrx{C}}$ is
\begin{equation}
  \tilde{\matrx{C}} = \matrx{C} + \eye\,s^2
\end{equation}

We compute the fully marginalized likelihood (FML) that a given pair of stars
has the same 3-space velocity with a small velocity difference drawn from a
zero-mean Gaussian with variance $s^2$ (hypothesis 1, $\mathcal{L}_1$), and
the FML of the stars having different 3-space velocities (hypothesis 2,
$\mathcal{L}_2$).
We use the FML ratio $\mathcal{L}_1/\mathcal{L}_2$ as a scalar for selecting
candidate co-moving stars, as described below in more detail.
To compute these FMLs, the likelihood functions $L_i, L_j$ are marginalized
over true 3-space velocity and distance for each star in the pair $(i,j)$.
\begin{align}
  \mathcal{L}_1 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \bs{v} \,
    L_i(\bs{v}, d_i, s^2) \,
    L_j(\bs{v}, d_j, s^2) \,
    p(\bs{v}) \, p(d_i \given \varpi_i) \, p(d_j \given \varpi_j) \\
  \mathcal{L}_2 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \bs{v}_i \, \dd^3 \bs{v}_j \,
    L_i(\bs{v}_i, d_i, 0) \,
    L_j(\bs{v}_j, d_j, 0) \,
    p(\bs{v}_i) \, p(\bs{v}_j) \, p(d_i \given \varpi_i) \, p(d_j \given \varpi_j). \label{eq:hyp2}
\end{align}
The marginalization integral for hypothesis 2 can be split into the product of
two simpler integrals $\mathcal{L}_2 = Q_i \, Q_j$ where
\begin{equation}
  Q = \int \, \dd d \, \dd^3 \bs{v} \, L(\bs{v}, d, 0) \, p(\bs{v}) \, p(d\given\varpi)
\end{equation}
If the velocity prior $p(\bs{v})$ is also Gaussian, the integrals over velocity
in both cases can be performed analytically:
\todo{SMOH: is this what we do?}
We use an equal-weight mixture of three isotropic, zero-mean Gaussian
distributions
\begin{equation}
  p(\bs{v}) = \frac{1}{3} \, \sum_{m=1}^3 \, \mathcal{N}(0, \sigma_m^2)
\end{equation}
with velocity dispersions $(\sigma_1, \sigma_2, \sigma_3) = (8, 32, 128)~\kms$
meant to represent young disk stars, old disk stars, and halo stars.
We derive the relevant expressions in Appendix~\ref{sec:appendix}.
After marginalizing over velocity, the likelihood integrands only depend on
distance; we numerically compute the integrals over the true distances to each
star in a pair, $d_1,d_2$, using Monte Carlo integration with $K$ samples from
the distance posterior pdfs:
\begin{equation}
  \int \, \dd d \, \tilde{L}(d) \, p(d\given\varpi) \approx
    \frac{1}{K} \, \sum_k^K \, \tilde{L}(d_k)
\end{equation}
where $\tilde{L}(d)$ is the velocity-marginalized likelihood function.
Through experimentation, we have found that $K=128$ samples are sufficient for
estimating the above integrals for stars with a wide range in parallax
signal-to-noise.

@@

\section{Some mathematics relevant to the above} \label{sec:appendix}

\todo{APW will clean this up}

In what follows, all vectors are column vectors, unless we have transposed them.
We indicate whether a symbol represents a scalar, vector, or tensor
(matrix) by context, not typesetting.
Properties of the squared exponential:
\begin{eqnarray}
  \ln\left[\int\exp(-\frac{1}{2}\,\transp{[x-\nu]}\,\inv{A}\,[x-\nu] - \Delta)\,\dd x\right]
  &=& +\frac{1}{2}\ln ||2\pi\,A|| -\Delta
  \quad ,
\end{eqnarray}
where $x$ and $\nu$ are $D$-dimensional vectors, $A$ is a positive definite
matrix, $\Delta$ is a scalar, and the integral is implicitly over all
of $D$-dimensional $x$-space.
We will need to complete squares.
If we equate
\begin{eqnarray}
  \frac{1}{2}\,\transp{[x-\nu]}\,\inv{A}\,[x-\nu] + \Delta
  &=& \frac{1}{2}\,\transp{x}\,\inv{A}\,x + \transp{x}\,B\,b + C
  \quad ,
\end{eqnarray}
where $B\,b$ is a $d$-vector, and $C$ is a scalar, then we find
\begin{eqnarray}
  \nu &=& -A\,B\,b
  \\
  \Delta & = & C - \frac{1}{2}\,\transp{\nu}\,\inv{A}\,\nu
  \quad .
\end{eqnarray}

Now in this case, we want to write down the probability density for the data
given the distances and velocity (the likelihood), and the prior pdf for
the velocity, and their product.
The prior pdf for the velocity is Gaussian:
\begin{eqnarray}
  p(v) &=& \normal(v\given 0,V)
  \\
  \ln p(v) &=& -\frac{1}{2}\,\ln||2\pi\,V|| -\frac{1}{2}\,\transp{v}\,\inv{V}\,v
  \\
  V &=& \sigma_v^2\,I
  \quad ,
\end{eqnarray}
where $v$ is the three-vector velocity, and $V$ is an isotropic
variance tensor.
Similarly, the likelihood for the data is also a Gaussian.
Before we write that down, we will make some definitions.
We construct a velocity-space data vector $y$ as follows:
\begin{eqnarray}
  \transp{y} &\equiv& [(d_1\,\mu_{\alpha1}), (d_1\,\mu_{\delta1}),
                       (d_2\,\mu_{\alpha2}), (d_2\,\mu_{\delta2})] \quad \mbox{or}
  \\
  \transp{y} &\equiv& [(d_1\,\mu_{\alpha1}), (d_1\,\mu_{\delta1}), (v_{r1}),
                       (d_2\,\mu_{\alpha2}), (d_2\,\mu_{\delta2}), (v_{r2})]
  \quad ,
\end{eqnarray}
where we have given two versions, a 4-component and an 6-component
version (depending on whether there are measured radial velocities),
the subscripts 1, 2 refer to the first and second stars (arbitrarily
ordered) of the pair, we have multiplied the strict observables by
distances $d_1, d_2$, which we are permitted to do (because we are
\emph{conditioning} on $d_1, d_2$), and note the wacky minus-ones.
The vector $y$ is a column vector that we have transposed to make it
easy to write down.
Fundamentally, our H1 model (same velocity) is
\begin{eqnarray}
  y &=& M\,v + \mathrm{noise}
  \\
  \transp{M} &=& [\hat{\alpha_1}, \hat{\delta_1},
                  \hat{\alpha_2}, \hat{\delta_2}] \quad \mbox{or}
  \\
  \transp{M} &=& [\hat{\alpha}_1, \hat{\delta}_2, \hat{r}_1,
                  \hat{\alpha}_2, \hat{\delta}_2, \hat{r}_2]
  \quad,
\end{eqnarray}
where $M$ is a $6\times 3$ or $8\times 3$ design matrix, with zeros
in two rows, and unit vectors along the other rows.
The noise (in the $y$ vector) is drawn from a Gaussian with ($6\times 6$ or $8\times 8$)
covariance matrix $\Sigma$, constructed from the distances $d_1, d_2$ and
the \gaia-provided individual-object covariance matrices.
Given all these definitions, the likelihood function is
\begin{eqnarray}
  p(\data\given v,d_1,d_2) &=& d_1^2\,d_2^2\,\normal(y\given M\,v, \Sigma)
  \\
  \ln p(\data\given v,d_1,d_2) &=& 3\,\ln d_1 + 3\,\ln d_2
  -\frac{1}{2}\,\ln||2\pi\,\Sigma|| \nonumber \\ && \quad
  -\frac{1}{2}\,\transp{[y-M\,v]}\,\inv{\Sigma}\,[y-M\,v]
  \quad ,
\end{eqnarray}
where the factor of $d_1^2\,d_2^2$ converts to units of per \gaia\ data
from units of per $y$.

>>>>

% packages
\definecolor{cbblue}{HTML}{3182bd}
\usepackage{microtype}  % ALWAYS!
\usepackage{amsmath}
\hypersetup{backref,breaklinks,colorlinks,urlcolor=cbblue,linkcolor=cbblue,citecolor=black}

@@

% define macros for math
\newcommand{\given}{\,|\,}
\newcommand{\normal}{{\mathcal{N}}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\transp}[1]{{#1}^{\!\mathsf{T}}}
\newcommand{\inv}[1]{{#1}^{-1}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\vperp}{\bs{v}^\perp}
\newcommand{\propm}{\bs{\mu}}
\newcommand{\mat}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bs{#1}}
\newcommand{\kms}{\rm km~s^{-1}}
\newcommand{\msun}{{\rm M}_\odot}
\newcommand{\data}{\mathrm{data}}
\newcommand{\snr}{[S/N]_\varpi}
\newcommand{\eye}{\mathbb{I}}

@@

For each star with $\snr > 32$, we use the reported parallax and parallax
error to compute a distance corrected for the Lutz-Kelker bias
(\citealt{lutzkelker}):
\begin{equation}
  d_{\rm LK} = 1000 \, \left[\frac{\varpi}{2} \,
    \left(1 + \sqrt{1 - \frac{16}{\snr^2}} \right) \right]^{-1} \, {\rm pc}
\end{equation}
We use this distance and the sky positions of each source to compute 3D
Cartesian positions, $\boldsymbol{x}$, for all stars and find the nearest
neighbor to each source in 3-space using the KDTree implementation in
\project{scikit-learn} (\citealt{scikit-learn}). We remove duplicates from
this sample and are left with 29870 pairs of stars. For all pairs, we then
compute the tangential velocity difference
\begin{equation}
  |\Delta \vec{v}_{\rm t}| = d_{\rm LK} \,
                            \sqrt{(\mu_{\alpha,1}-\mu_{\alpha,2})^2 +
                                  (\mu_{\delta,1}-\mu_{\delta,2})^2}
\end{equation}
\figname~\ref{fig:dv-sep} shows the tangential velocity difference against the
physical separation plotted for all pairs that pass the above signal-to-noise
cut. Pairs with small physical separation also tend to have small velocity
differences. Intriguingly, there appear to be an abundance of pairs with small
velocity separation at separations $|\delta\vec{x}| \gtrsim 1~{\rm pc}$. Of
course, this figure really only highlights close pairs with small angular
separations:
At large angular separations, the tangential velocity differences will be large
because of projection effects.
To select a more representative population of co-moving stars, we must instead
model the 3-space velocities of the stars given observations of just two
components.

@@

The abundance of pairs of stars with small velocity difference in the
high-precsision sample of \figname~\ref{fig:dv-sep} suggests that there are a
significant number of co-moving stars in the \tgas\ data at separations as large
as a few parsecs. Here we develop a method to select high-confidence co-moving
stars that properly incorporates the complex uncertainties associated with the
\gaia\ data. We make the following assumptions in order to construct a
statistical model (a likelihood function with explicit priors on our
parameters):
\begin{itemize}
  \item We assume that the uncertainties in the data---parallax, $\varpi$, and
    two proper motion components, $\propm = (\begin{array}[t]{c c} \mu_\alpha &
    \mu_\delta\end{array})^\mathsf{T}$---are Gaussian with known covariances
    $\mat{C}$.
  \item We assume that the 3-space velocities of a given pair of stars
    $(\vec{v}_i, \vec{v}_j)$ in the \tgas\ sample (relative to the solar system
    barycenter) are either (1) co-moving with velocity drawn from a velocity
    prior $p(\vec{v})$ and  velocity difference drawn from a zero-mean Gaussian
    with velocity dispersion $s$, or (2) individually drawn from a velocity
    prior $p(\vec{v})$.
\end{itemize}

Under these assumptions, the likelihood of a proper motion measurement for a
star given true distance, $d$, and true tangential velocity, $\vec{v}^\perp =
(\begin{array}[t]{c c} v_\alpha & v_\delta\end{array})^\mathsf{T}$ is
\begin{align}
  L(\vec{v}, d, s^2) &=
    \left[\det\left(\frac{\tilde{\mat{C}}^{-1}}{2\pi}\right)\right]^{1/2} \,
    \exp \left[ -\frac{1}{2} \transp{\left(\vec{\mu} - \vec{x}_\theta \right)} \,
    \tilde{\mat{C}}^{-1} \,
    \left(\vec{\mu} - \vec{x}_\theta \right) \right] \label{eq:likefn} \\
  \vec{x}_\theta &= d^{-1} \, \vec{v}^\perp
\end{align}
where the tangential velocity $\vec{v}^\perp$ is related to the 3-space velocity
$\vec{v}$ through projection onto a tangent plane at a given star's sky position
$(\alpha, \delta)$
\begin{align}
  \vec{v}^\perp &= \mat{M}\,\vec{v} \\
  & = \left(
      \begin{array}{c c c}
        -\sin\alpha & \cos\alpha & 0 \\
        -\sin\delta \, \cos\alpha & -\sin\delta \, \sin\alpha & \cos\delta
      \end{array}
    \right) \,
    \left(\begin{array}{c} v_x \\ v_y \\ v_z \end{array}\right)
\end{align}
and the modified covariance matrix $\tilde{\mat{C}}$ is
\begin{equation}
  \tilde{\mat{C}} = \mat{C} + s^2 \, \eye
\end{equation}

We compute the fully marginalized likelihood (FML) that a given pair of stars
has the same 3-space velocity with a small velocity difference drawn from a
zero-mean Gaussian with variance $s^2$ (hypothesis 1, $\mathcal{L}_1$), and
the FML of the stars having different 3-space velocities (hypothesis 2,
$\mathcal{L}_2$).
We use the FML ratio $\mathcal{L}_1/\mathcal{L}_2$ as a scalar for selecting
candidate co-moving stars, as described below in more detail.
To compute these FMLs, the likelihood functions $L_i, L_j$ are marginalized
over true 3-space velocity and distance for each star in the pair $(i,j)$.
\begin{align}
  \mathcal{L}_1 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \vec{v} \,
    L_i(\vec{v}, d_i, s^2) \,
    L_j(\vec{v}, d_j, s^2) \,
    p(\vec{v}) \, p(d_i \given \varpi_i) \, p(d_j \given \varpi_j) \\
  \mathcal{L}_2 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \vec{v}_i \, \dd^3 \vec{v}_j \,
    L_i(\vec{v}_i, d_i, 0) \,
    L_j(\vec{v}_j, d_j, 0) \,
    p(\vec{v}_i) \, p(\vec{v}_j) \, p(d_i \given \varpi_i) \, p(d_j \given \varpi_j). \label{eq:hyp2}
\end{align}
The marginalization integral for hypothesis 2 can be split into the product of
two simpler integrals $\mathcal{L}_2 = Q_i \, Q_j$ where
\begin{equation}
  Q = \int \, \dd d \, \dd^3 \vec{v} \, L(\vec{v}, d, 0) \, p(\vec{v}) \, p(d\given\varpi)
\end{equation}
If the velocity prior $p(\vec{v})$ is also Gaussian, the integrals over velocity
in both cases can be performed analytically:
\todo{SMOH: is this what we do?}
We use an equal-weight mixture of three isotropic, zero-mean Gaussian
distributions
\begin{equation}
  p(\vec{v}) = \frac{1}{3} \, \sum_{m=1}^3 \, \mathcal{N}(0, \sigma_{v,m}^2)
\end{equation}
with velocity dispersions $(\sigma_{v,1}, \sigma_{v,2}, \sigma_{v,3}) = (8, 32,
128)~\kms$ meant to represent young disk stars, old disk stars, and halo stars.
We derive the relevant expressions in Appendix~\ref{sec:appendix}.
After marginalizing over velocity, the likelihood integrands only depend on
distance; we numerically compute the integrals over the true distances to each
star in a pair, $d_1,d_2$, using Monte Carlo integration with $K$ samples from
the distance posterior pdfs:
\begin{equation}
  \int \, \dd d \, \tilde{L}(d) \, p(d\given\varpi) \approx
    \frac{1}{K} \, \sum_k^K \, \tilde{L}(d_k)
\end{equation}
where $\tilde{L}(d)$ is the velocity-marginalized likelihood function.
Through experimentation, we have found that $K=128$ samples are sufficient for
estimating the above integrals for stars with a wide range in parallax
signal-to-noise.

\todo{APW/SMOH: in results, specify what we set $s^2$ to:}
To search for wide binaries, we set $s^2 = \frac{2 \, G \, \msun}{|\vec{x}_i-\vec{x}_j|}$.
To search for dissolving open clusters, we set $s^2 = 1~(\kms)^2$.

@@

\section{Relevant properties of Gaussian integrals}
\label{sec:appendixA}

In what follows, all vectors are column vectors, unless we have transposed them.
A relevant exponential integral solution is
\begin{eqnarray}
  \ln\left[\int\exp(-\frac{1}{2}\,
    \transp{[\vec{x}-\vec{\nu}]} \,
    \inv{\mat{A}} \,
    [\vec{x}-\vec{\nu}] - \Delta) \, \dd \vec{x}\right]
  &=& +\frac{1}{2}\ln ||2\pi\,\mat{A}|| -\Delta
  \quad , \label{eq:gauss-int}
\end{eqnarray}
where $\vec{x}$ and $\vec{\nu}$ are $D$-dimensional vectors, $\mat{A}$ is a
positive definite matrix, $\Delta$ is a scalar, and the integral is over all of
$D$-dimensional $\vec{x}$-space.
To cast our problem in this form, we will need to complete the square of the
exponential argument.
If we equate
\begin{eqnarray}
  \frac{1}{2}\,\transp{[\vec{x}-\vec{\nu}]}\,\inv{\mat{A}}\,[\vec{x}-\vec{\nu}] + \Delta
  &=& \frac{1}{2}\,\transp{\vec{x}}\,\inv{\mat{A}}\,\vec{x} + \transp{\vec{x}}\,\mat{B}\,\vec{b} + C
  \quad ,
\end{eqnarray}
where $\mat{B}\,\vec{b}$ is an $D$-vector, and $C$ is a scalar, then we find
\begin{eqnarray}
  \vec{\nu} &=& -\mat{A}\,\mat{B}\,\vec{b}
  \\
  \Delta & = & C - \frac{1}{2}\,\transp{\vec{\nu}}\,\inv{\mat{A}}\,\vec{\nu}
  \quad .
\end{eqnarray}
We will identify terms in our likelihood functions with $\mat{A}$,
$\mat{B}\,\vec{b}$, and $C$, convert to $\vec{\nu}$ and $\Delta$ and compute
the marginalized likelihood using \eqname~\ref{eq:gauss-int}.

\section{Expressions for the marginalized likelihoods}\label{sec:appendix}

At given distance $d$, the velocity-marginalized likelihood can be computed
analytically using the expressions in Appendix~\ref{sec:appendixA}.
We will start by writing down expressions for the the likelihood multiplied by
the prior pdf for the velocities.
As described above, we use an isotropic, mixture-of-Gaussians prior on velocity
where the velocity dispersion of a given mixture component is $\sigma_{v,m}$ and
the 3-space variance tensor is $\mat{V}_m = \sigma_{v,m}^2 \, \eye$.
The likelihood for the data is also a Gaussian, as shown in
\eqname~\ref{eq:likefn}.

Similarly, the likelihood for the data is also a Gaussian.
Before we write that down, we will make some definitions.
We construct a velocity-space data vector $y$ as follows:
\begin{eqnarray}
  \transp{y} &\equiv& [(d_1\,\mu_{\alpha1}), (d_1\,\mu_{\delta1}),
                       (d_2\,\mu_{\alpha2}), (d_2\,\mu_{\delta2})] \quad \mbox{or}
  \\
  \transp{y} &\equiv& [(d_1\,\mu_{\alpha1}), (d_1\,\mu_{\delta1}), (v_{r1}),
                       (d_2\,\mu_{\alpha2}), (d_2\,\mu_{\delta2}), (v_{r2})]
  \quad ,
\end{eqnarray}
where we have given two versions, a 4-component and an 6-component
version (depending on whether there are measured radial velocities),
the subscripts 1, 2 refer to the first and second stars (arbitrarily
ordered) of the pair, we have multiplied the strict observables by
distances $d_1, d_2$, which we are permitted to do (because we are
\emph{conditioning} on $d_1, d_2$), and note the wacky minus-ones.
The vector $y$ is a column vector that we have transposed to make it
easy to write down.
Fundamentally, our H1 model (same velocity) is
\begin{eqnarray}
  y &=& M\,v + \mathrm{noise}
  \\
  \transp{M} &=& [\hat{\alpha_1}, \hat{\delta_1},
                  \hat{\alpha_2}, \hat{\delta_2}] \quad \mbox{or}
  \\
  \transp{M} &=& [\hat{\alpha}_1, \hat{\delta}_2, \hat{r}_1,
                  \hat{\alpha}_2, \hat{\delta}_2, \hat{r}_2]
  \quad,
\end{eqnarray}
where $M$ is a $6\times 3$ or $8\times 3$ design matrix, with zeros
in two rows, and unit vectors along the other rows.
The noise (in the $y$ vector) is drawn from a Gaussian with ($6\times 6$ or $8\times 8$)
covariance matrix $\Sigma$, constructed from the distances $d_1, d_2$ and
the \gaia-provided individual-object covariance matrices.
Given all these definitions, the likelihood function is
\begin{eqnarray}
  p(\data\given v,d_1,d_2) &=& d_1^2\,d_2^2\,\normal(y\given M\,v, \Sigma)
  \\
  \ln p(\data\given v,d_1,d_2) &=& 3\,\ln d_1 + 3\,\ln d_2
  -\frac{1}{2}\,\ln||2\pi\,\Sigma|| \nonumber \\ && \quad
  -\frac{1}{2}\,\transp{[y-M\,v]}\,\inv{\Sigma}\,[y-M\,v]
  \quad ,
\end{eqnarray}
where the factor of $d_1^2\,d_2^2$ converts to units of per \gaia\ data
from units of per $y$.
